A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-13 Simulation/Writeup/NoteConvergence.Rnw
results

There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}. Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}. Algorithmic convergence should, therefore, be monitored carefully. In that respect is the current evaluation practice insuffient. That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}. Quantitative evaluation of the convergence of MI algorithms would be preferred.

@@

Loosely speaking, the aim of MCMC algorithms is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in an MI algorithm). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending across iterations within chains. %\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

\textbf{[COULD BE SHORTER!]} Most convergence diagnostics target either of the two components. The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains. The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}. With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
Other convergence diagnostics are outside the scope of this study.\footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and autocorrelation.

@@

%The potential scale reduction factor $\widehat{R}$ is obtained as:
We can interpret $\widehat{R}$ as potential scale reduction factor, since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}. High $\widehat{R}$-values would therefore indicate slow mixing of chains, e.g., because of dependence across chains on initial values. Once there is no dependence on initial values anymore, there is no added value of continuing iterations. \textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$). 
%If there is no added value compared to infinitely many iterations, the chains have converged. 
The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

Following the same notation, autocorrelation is defined as the correlation between two subsequent $\theta$ values within the same chain. We only consider $AC$ at lag 1, i.e., between the $t^{th}$ and $t+1^{th}$ iteration of each chain: 
%
\begin{equation*}
AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2},
\end{equation*}
%
where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}. We can interpret the sign of $AC$-values as measure of dependence within chains ($-1 \leq AC \leq 1$). Positive $AC$-values would indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending. Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. As convergence diagnostic, the interest is in positive $AC$-values.\footnotemark 
\footnotetext{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities (e.g., bias in estimating $Q$).

@@

%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
We use model-based simulation to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. Convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage) are computed for each simulation condition, and aggregated across simulation runs.
%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition. %All simulation conditions are replicated 1000 times. %For each of the 100 conditions 1000 simulation runs are performed. for each of the 100 simulation conditions. \textbf{Estimands are convergence diagnostics. Performance measures are simulation diagnostics.  where for one through one hundred the Rhat and ... are calculated.} 
The simulation set-up consists of several steps, summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available on Github.}

@@

\subsection{Data simulation}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1 \text{ and } Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$. The quantity of scientific interest is regression coefficient $\beta_1$. The data generating model is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.

@@

\subsection{Amputation}
The complete data is the point of origin for all simulations. In each repetition of the simulation study, the complete data is `amputed' once. That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\%. This leaves 20\% of the rows completely observed. The resulting amputed data is equal for all simulation conditions in the same repetition.%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

\subsection{Imputation}
Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$). The number of iterations is varied over simulations (\texttt{maxit} argument between 1 and 100). Each simulation condition (i.e., each number of iterations) is simulated 1000 times. Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.

$\widehat{R}$ is computed within imputation chains: for each variable, each simulation condition, and each simulation. The maximum value across variables within the same simulation is reported. Maximum $\widehat{R}$-values per simulation condition are aggregated across simulations.

autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

Linear regression is performed using the \proglang{R} function \fct{lm}. The quantity of scientific interest $Q$ is the regression coefficient of $X$ on $Y$. $Q$ is estimated in each imputation, in each simulation condition, in each repetition. The $m = 5$ estimated regression coefficients $\hat{Q}$ are pooled according to Rubin's \citeyear{rubin87} rules with \pkg{mice} function \fct{pool}. Variance is computed accroding to finite population pooling \textbf{Vink and Van Buuren, 2014??}.

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \citep[\S~2.5.2]{buur18}. These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data. Under anything but the MCAR assumption, this can be expected. When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations). When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.

% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value

Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.

Confidence interval width (CIW) is computed as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%). The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom). CIW is averaged over all simulations with the same simulation condition.

Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.

@@

Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions. A subset of simulation conditions is presented in Table \ref{tab:results}. This section describes convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.

@@

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
It. & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}

\textbf{$\widehat{R}$}. The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct! Also say something about Rh below 1.}

\textbf{autocorrelation.} All observed autocorrelations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} 
at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

>>>>

There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}. Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}. Algorithmic convergence should, therefore, be monitored carefully. In that respect is the current evaluation practice insufficient. That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}. Quantitative evaluation of the convergence of MI algorithms would be preferred.

@@

Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending across iterations within chains. %\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

\textbf{[COULD BE SHORTER!]} Most convergence diagnostics target either of the two components. The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains. The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}. With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
Other convergence diagnostics are outside the scope of this study.\footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and autocorrelation.

@@

%The potential scale reduction factor $\widehat{R}$ is obtained as:
We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}. High $\widehat{R}$-values, therefore, indicate slow mixing of chains, e.g., because of dependence across chains on initial values. Once there is no dependence on initial values anymore, there is no added value of continuing iterations. \textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$). 
%If there is no added value compared to infinitely many iterations, the chains have converged. 
The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

Following the same notation, autocorrelation is defined as the correlation between two subsequent $\theta$ values within the same chain. We only consider $AC$ at lag 1, i.e., between the $t^{th}$ and $t+1^{th}$ iteration of each chain: 
%
\begin{equation*}
AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2},
\end{equation*}
%
where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}. We can interpret the sign of $AC$-values as a measure of dependence within chains ($-1 \leq AC \leq 1$). Positive $AC$-values indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending. Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. As convergence diagnostic, the interest is in positive $AC$-values.\footnotemark 
\footnotetext{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities. The perfrormance diagnostics are the evaluation criteria  recommended by \citep[\S~2.5.2]{buur18}, whcih comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage). These diagnostics are aggregated across repetitions to obtain averages per simulation condition.
%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition. %All simulation conditions are replicated 1000 times. %For each of the 100 conditions 1000 simulation runs are performed. for each of the 100 simulation conditions. \textbf{Estimands are convergence diagnostics. Performance measures are simulation diagnostics.  where for one through one hundred the Rhat and ... are calculated.} 
The simulation set-up consists of several steps, summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available on Github.}

@@

\subsection{Data simulation}
The point of origin for all simulation conditions is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$. The quantity of scientific interest is regression coefficient $\beta_1$. The data generating model is a multivariate normal distribution with means structure $\mu$ and variance-covariance matrix $\Sigma$.

@@

\subsection{Amputation}
The complete dataset is `amputed' once in each simulation repetition. That is, the \pkg{mice} function \fct{ampute} is used to impose a missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}). This leaves 20\% of the rows completely observed. %The resulting amputed data is equal for all simulation conditions in the same repetition. %"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

\subsection{Imputation}
Missing data points are imputed with the \pkg{mice} function \fct{mice}. All simulations are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}). The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}). 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
We investigate convergence by extracting not only the imputed values of each MI procedure (the $T^{th}$ iteration), but also information ($\theta$-values) from the intermediate iterations ($t = 1, 2, \dots, T$). The $\theta$s of interest are chain means and chain variances per variable. We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing the definition given in \textbf{EQUATION NR}. $AC$ is computed with \pkg{stats} function \fct{acf}.

% 
% 
% extract chain means and chain variances per iteration $t$. From the MI algorithm Convergence diagnostics are computed for chain means and chain variances.   
% $\widehat{R}$ is computed within imputation chains: . The maximum value across variables within the same simulation is reported. 
% 
% autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.

@@

To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the \pkg{stats} function \fct{lm}. We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$. We use the \pkg{mice} function \fct{pool} to get variance estimates according to Rubin's \citeyear{rubin87} rules, and utilize these to implement finite population pooling as per \citep{vink14}.

\subsection{Simulation diagnostics}
% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data. Under anything but the MCAR assumption, this can be expected. When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations). When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study. 
% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value

We compute bias as the difference between $\bar{Q}$ and $Q$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%). We compute the CI95\% bounds as $\bar{Q} \pm 2.66 \times SE_{\bar{Q}}$, where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

Figures \ref{fig:conv} and \ref{fig:sim} display results per simulation condition ($T = 1,2,\dots,100$). A subset of simulation conditions is presented in Table \ref{tab:results}.

@@

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
T & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$). Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero). The convergence diagnostics show more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.

%The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
\textbf{say something about Rh below 1.}
%%%%

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. At $T=5$, the $AC$-value reaches the level observed at $T=2$. Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$. $AC$ is then indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

% All observed autocorrelations  are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} 
% at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

Taken together, we see that convergence may be diagnosed in simulation conditions where $T>10$.

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average condidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).
